---
title: "Bob's Burgers Episodes"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "light"
    downcute_theme: "default"
---


```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE)
```

## Load R Packages used

```{r}
pacman::p_load(
  ggplot2,
  reshape2,
  png,
  grid,
  viridis,
  here,
  tidylog
)
```

## The Data

```{r}
# Option 1: tidytuesdayR package 
## install.packages("tidytuesdayR")
tuesdata <- tidytuesdayR::tt_load(2024, week = 47)

episode_metrics <- tuesdata$episode_metrics

```



## Finding Patterns

```{r eval=FALSE}
# Prepare the data
heatmap_data <- dcast(episode_metrics, episode ~ season, value.var = "dialogue_density")

# Melt data for ggplot
heatmap_long <- melt(heatmap_data, id.vars = "episode", variable.name = "season", value.name = "dialogue_density")
```



```{r fig.width=8, fig.height=6}
# Load and desaturate the Bob's Burgers image
img <- readPNG("bobs_burger_image.png")  # Replace with your image path
background_image <- rasterGrob(img, width = unit(1, "npc"), height = unit(1, "npc"), just = "center")

# Create the heatmap
(plot <- ggplot(data = heatmap_long, aes(x = as.factor(season), y = as.factor(episode), fill = dialogue_density)) +
  annotation_custom(background_image, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +  # Add image as background
  geom_tile(color = "white", size = 0.3, alpha = 0.9) +  # Add tile borders and set transparency
  scale_fill_viridis(
    option = "viridis", 
    direction = -1, 
    na.value = "grey50", 
    name = "Dialogue Density\n(Grey = Missing Episodes)",
    guide = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = 10,
      barheight = 1
    )
  ) +
  labs(
    title = "Dialogue Density by Episode and Season",
    x = "Season",
    y = "Episode",
    caption = "#TidyTuesday Wk 47: Bob's Burgers Episodes\n @njokinjuki"
  ) +
  theme_minimal(base_family = "Arial") +  # Use a clean font
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold", color = "black"),
    axis.text = element_text(size = 10, color = "black"),
    axis.title = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8),
    legend.position = "bottom",  # Place the legend at the bottom
    panel.background = element_rect(fill = "transparent", color = NA),  # Transparent background
    plot.background = element_rect(fill = "white", color = NA),         # White outer border
    axis.ticks = element_blank(),  # Clean axes
    panel.grid = element_blank()   # Remove default gridlines
  ))
```

```{r}
# save the plot
ggsave(plot = plot, file = here::here("Plots", "Wk47-Bob_Burgers_Episodes.png"), width = 6, height = 6, bg = "white")
```


## Interpretion of the plot

The heatmap shows dialogue density by episode (y-axis) and season (x-axis). 

- High Dialogue Density (Purple): Episodes or seasons with darker purple shades have higher dialogue density. This means characters talk more or the episodes have more spoken content.

- Low Dialogue Density (Yellow Shades): Episodes or seasons with lighter shades (yellow) have less dialogue density, possibly indicating episodes with more action, visual gags, or silence.

Trends Across Seasons:

- If earlier seasons (left) have consistent red patterns, the show may have started with a strong focus on dialogue-heavy humor.
- Variations or patterns in specific seasons (e.g., alternating colors) may indicate stylistic changes in how dialogue is used.

Episode-Specific Insights:

- Episodes with higher or lower densities might correlate with key moments in the show, such as plot-heavy episodes, musical numbers, or experimental episodes.





