---
title: "Wk 36: Stack Overflow Annual Developer Survey 2024"
subtitle: "TidyTuesday Contribution"
author: "Njoki Njuki"
date: "2024-09-03"
date-modified: today
format: html
toc: true
code-fold: show
editor: visual
---

## Load Packages

```{r}

pacman::p_load(
  tidyverse, # for data transformation and plots
  ggthemr,   # for theme
  ggtext     # for texts
  
)
```

## Data

This week's dataset comes from the 2024 Stack Overflow Annual Developer Survey, conducted in May 2024. The survey received responses from over 65,000 developers, covering seven main areas:

-   Basic Information
-   Education, Work, and Career
-   Tech and Tech Culture
-   Stack Overflow Community
-   Artificial Intelligence (AI)
-   Professional Developer Series (separate from the main survey)
-   Thoughts on the Survey


The dataset for analysis includes only single-response questions from the main sections, with categorical responses coded as integers. Labels for these codes can be found in the crosswalk file. Key points to explore include developer demographics, their engagement with Stack Overflow, and their views on AI.

```{r}
#| label: import_data
#| message: false
#| warning: false

qname_levels_single_response_crosswalk <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-09-03/qname_levels_single_response_crosswalk.csv')
stackoverflow_survey_questions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-09-03/stackoverflow_survey_questions.csv')
stackoverflow_survey_single_response <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-09-03/stackoverflow_survey_single_response.csv')

```

## Data preparation

```{r}
#| label: data_prep1
#| message: false
#| warning: false

# Ensure qname_levels_single_response_crosswalk has the proper columns and data
# Filter only for rows where qname is 'dev_type' and create a mapping
dev_type_mapping <- qname_levels_single_response_crosswalk %>%
  filter(qname == "dev_type") %>%
  select(level, label) |> 
  mutate(level = as.character(level))

# Rename dev_type levels in the original dataset using the mapping
stackoverflow_survey_single_response.1 <- stackoverflow_survey_single_response %>%
  # Convert dev_type to character to match the type of the 'label' column
  mutate(dev_type = as.character(dev_type)) %>%
  left_join(dev_type_mapping, by = c("dev_type" = "level")) %>%
  mutate(dev_type = coalesce(label, dev_type)) %>%  # Replace levels with labels
  select(-label)  # Drop the label column after replacement

# Select top 20 developer type
dev_data <- stackoverflow_survey_single_response.1 %>%
  group_by(dev_type) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  na.omit() %>%
  slice(1:20) %>%
  ungroup()
```



```{r}
#| label: data_prep2
#| message: false
#| warning: false

# Filter the original data for only the top 20 dev_types
top_20_data <- stackoverflow_survey_single_response.1 %>%
  filter(dev_type %in% dev_data$dev_type) %>%
  mutate(ed_level = case_when(
    ed_level == 1 ~ "Associate degree",
    ed_level == 2 ~ "Bachelor’s degree",
    ed_level == 3 ~ "Master’s degree",
    ed_level == 4 ~ "Primary/elementary school",
    ed_level == 5 ~ "Professional degree",
    ed_level == 6 ~ "Secondary school",
    ed_level == 7 ~ "Some college/uni study w/o earning a degree",
    ed_level == 8 ~ "Something else"
  )) 

# Group by dev_type and ed_level, then count how many in each group
ed_level_distribution <- top_20_data |> 
  group_by(dev_type, ed_level) |> 
  summarise(count = n()) |> 
  ungroup()
```

## Plot 

```{r}
#| label: create_plot
#| message: false
#| warning: false
#| fig-width: 8.5
#| fig-height: 7

# Setting the theme
ggthemr("pale")

# Create the plot
(p <- ggplot(ed_level_distribution) +     
  
  # Create the bar chart
  geom_bar(stat = "identity",
           aes(x = reorder(dev_type, count, desc = TRUE), 
               y = count,
               fill = ed_level)) +
  
  # Add title, x and y axis labels, and the legend label
  labs(
    title = "Education Level Distribution by <span style='color:maroon;'>Top 20 Developer Types</span>",
    subtitle = "2024 Stack Overflow Annual Developer Survey",
    x = "Developer Type",
    y = "Frequency",
    fill = "Education Level",
    caption = "#TidyTuesday Wk 36: Stack Overflow Annual Developer Survey 2024; @njokinjuki"  # Add caption
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  # Center the title, place legend at the bottom, and format the title with color
  theme_minimal() +
  theme(
    plot.title = element_markdown(hjust = 0, size = 12),  
    plot.subtitle = element_markdown(size = 10),
    legend.position = "bottom",                  # Move legend to the bottom
    plot.caption = element_text(hjust = 0.5)     # Center the caption
  ) +
   # Flip coordinates to make it horizontal
  coord_flip() +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE, label.hjust = 0, 
                               label.position = "bottom")) +
  # Use the Rick and Morty color palette
  ggsci::scale_fill_futurama(name = "Education Level")+
    
    annotate("text",
         x = 4, y = max(ed_level_distribution$count) * 0.99,  # Position text near the top of the plot
         label = expression(italic("Full Stack Developers responded the most, \nwith the majority holding a Bachelor's degree") * " 💻"),  # Italic and computer icon
         size = 3.5, hjust = 0, color = "#525252")


)

# save the plot
# ggsave(plot = p, file = here::here("September","Plots", "2024_Stack_Overflow_Annual_Developer_Survey.png"), width = 9, height = 8.5, bg = "white")

# turn off the theme effect!
ggthemr_reset()
```

## Pun for the week!

Why don't statisticians argue with R?

Because it always **make a good point!** `r emoji::emoji("smile")` `r emoji::emoji("bar_chart")`
